---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```
# pensionfinanceLi

<!-- badges: start -->
[![Travis build status](https://travis-ci.org/sstoeckl/pensionfinanceLi.svg?branch=master)](https://travis-ci.org/sstoeckl/pensionfinanceLi)
[![AppVeyor build status](https://ci.appveyor.com/api/projects/status/github/sstoeckl/pensionfinanceLi?branch=master&svg=true)](https://ci.appveyor.com/project/sstoeckl/pensionfinanceLi)
[![Codecov test coverage](https://codecov.io/gh/sstoeckl/pensionfinanceLi/branch/master/graph/badge.svg)](https://codecov.io/gh/sstoeckl/pensionfinanceLi?branch=master)
<!-- badges: end -->

The goal of pensionfinanceLi is to optimize the situation of a typical person insured in Liechtenstein's Pension System. Currently, however, we do not implement any special cases. For a detailed documentation check `vignette("model")`.

## Installation

<!-- You can install the released version of pensionfinanceLi from [CRAN](https://CRAN.R-project.org) with: -->

<!-- ```{r instpkg, eval=FALSE} -->
<!-- install.packages("pensionfinanceLi") -->
<!-- ``` -->

Currently this package is under development, but can be used subject to careful control. It can be installed from [github](https://github.com) or [BitBucket](https://bitbucket.org/) with:

```{r instgit, eval=FALSE}
# install.packages("devtools")
devtools::install_github("sstoeckl/pensionfinanceLi")
devtools::install_bitbucket("sstoeckl/pensionfinanceLi")
```
